---
title: "Fear index"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Fear index}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, 
                      message = FALSE, 
                      warning = FALSE, 
                      collapse = FALSE,
                      comment = "#>",
                      fig.align = "center")
knitr::opts_chunk$set(fig.width = 12, fig.height = 9)
set.seed(1)
extrafont::loadfonts(quiet=TRUE)
options(scipen = 999) # turn-off scientific notation like 1e+48
library(unhcRstyle)
library(ggplot2)
library(mice)
library(plyr) 
library(GGally) 

```



The following variables related to fear and violence were measured in all three Northern Triangle countries:

- fear10: Has avoided walking through dangerous areas
- vic44: Organized in Neighborhood for Security (0 = no, 1=yes) 
- vic1exta: Victim of Crime (Frequency) (999999 if vic1ext=2, otherwise max=15) 
- vic1hogar: Other Victim of Crime in Household (1=yes, 2=no)
- aoj11: Perception of Neighborhood Security (1 = safe, 4 = unsafe) 
- pese1: Perception of Neighborhood Violence (1=high, 3=low) 
- pese2: Trend in Neighborhood Violence (1=high, 3=low) 
- aoj17: Gang Presence in Neighborhood (1 = a lot, 4 = none) 
- diso7: Youth Loitering a Problem in the Neighborhood (1 = very serious, 5 = not a problem) 
- diso8: Youth in Gangs a Problem in the Neighborhood (1 = very serious, 5 = not a problem) 
- diso10: Illegal Drug Trafficking a Problem in the Neighborhood (1 = very serious, 5 = not a problem) 
- diso18: Gang Fights a Problem in the Neighborhood (1 = very serious, 5 = not a problem) 
- diso14: Drug Addicts a Problem in the Neighborhood (1 = very serious, 5 = not a problem) 
- diso16: Assaults a Problem in the Neighborhood (1 = very serious, 5 = not a problem) 
- diso17: Shootings a Problem in the Neighborhood (1 = very serious, 5 = not a problem) 
- vicbar1: Burglaries in the Neighborhood (1=yes, 2=no) 
- vicbar1f: Number of Occurrences of Burglaries in Neighborhood (1 = weekly, 3=yearly) - vicbar3: Sales of Illegal Drugs in Neighborhood (1=yes, 2=no) 
- vicbar4: Extortion or Blackmail in the Neighborhood (1=yes, 2=no)
- vicbar7: Murders in the Neighborhood (1=yes, 2=no) 
- a4 == 5,14,27,30,57: Biggest problem related to crime/violence/security

`vic44` appears to be on-topic, but actually doesn't correlate as strongly as the other variables do, so we're leaving it out.

```{r,message=FALSE}
 
mainDir <- getwd()
## If you save your analysis under vignette folder...
mainDirroot <- substring(mainDir, 0 , nchar(mainDir) - 10)


lapop.2014.GTM <- read.csv(paste0(mainDirroot, "/data-raw/lapop.2014.GTM.csv"))
lapop.2014.SLV <- read.csv(paste0(mainDirroot, "/data-raw/lapop.2014.SLV.csv"))
lapop.2014.HND <- read.csv(paste0(mainDirroot, "/data-raw/lapop.2014.HND.csv"))

v <- c('fear10','vic1ext','vic1exta','vic1hogar','aoj11','pese1','pese2',
       'aoj17','diso7','diso8','diso10','diso18','diso14','diso16','diso17',
       'vicbar1','vicbar1f','vicbar3','vicbar4','vicbar7','a4')
my_data <- rbind(lapop.2014.GTM[,v],lapop.2014.SLV[,v],lapop.2014.HND[,v])

my_data[my_data$vic1exta > 800000,'vic1exta'] <- 0

my_data$a4 <- as.numeric(my_data$a4 %in% c(5,14,27,30,57))

is.na(my_data[my_data>800000]) <- TRUE

my_data <- my_data[,order(names(my_data))] # alphabetize columns

my_complete <- na.omit(my_data)
pr_complete <- prcomp(my_complete, center = TRUE, scale = FALSE)
plot(pr_complete)
summary(pr_complete)
```

41.3% of total variance winds up in the first component.

```{r}
my_imp <- mice(my_data, printFlag = F,seed=12345)
nmis <- my_imp$nmis
nm <- data.frame(nmis=nmis/nrow(my_data),var=names(nmis))
ggplot(nm,aes(x=var,y=nmis)) +
  geom_bar(stat = 'identity', fill = '#90AFC5') +
  ylab('Fraction of imputed values') +
  coord_flip() +
  theme_classic() +
  theme(legend.position='none',
        axis.ticks=element_blank(),
        axis.line=element_blank(),
        axis.title.y=element_blank(),
        text=element_text(size=20))
```

The only variable that needed a lot of imputation was `vicbar1f` -- the number of burglaries in the neighborhood. I'm guessing this is because people don't always know about all the burglaries that might be going on. Let's look at the PCA for one of our imputed datasets.

```{r}
pr <- lapply(1:5,function(x) prcomp(complete(my_imp,x),scale=FALSE,center=TRUE))
plot(pr[[1]])
summary(pr[[1]])
```

Imputation decreased the fraction of variance in the first component to 46.7%.

```{r}
pc1 <- pr[[1]]$x[,1]
pc2 <- pr[[1]]$x[,2]
qplot(pc1,pc2) + theme_classic()
```

The distribution along PC1 looks pretty continuous, while the PC2 distribution looks very skewed. Fortunately, we're only using PC1.

```{r}
all_pc1 <- data.frame(llply(1:5, function(i) pr[[i]]$x[,1]))
names(all_pc1) <- c('imp1', 'imp2', 'imp3', 'imp4', 'imp5')
ggpairs(all_pc1) + theme_classic()
all_pc1$avg <- rowMeans(all_pc1)
```

Correlations are high, and the distribution is skewed toward high values.

```{r,fig.height=8,fig.width=10}
qchange <- data.frame(low=colMeans(my_data[all_pc1$avg < quantile(all_pc1$avg)[2],],
                                   na.rm=TRUE),
                      hi=colMeans(my_data[all_pc1$avg > quantile(all_pc1$avg)[4],],
                                  na.rm=TRUE))
qchange$slope <- qchange$hi - qchange$low
qchange$label <- rownames(qchange)
qchange$even <- rank(qchange$hi)/nrow(qchange)*max(qchange$hi)
ggplot(data=qchange) +
  geom_segment(aes(x=0,y=qchange$low,xend=1,yend=qchange$hi,
                   color=qchange$slope),size=1) +
  scale_color_gradient(low='#FF420E',high='#89DA59') +
  geom_text(aes(x=1,y=(qchange$even+qchange$hi)/2,label=qchange$label,
                hjust=0,color=qchange$slope),size=5) +
  annotate("text",x=0,y=0,label='Average in lowest quartile',hjust=0) +
  annotate("text",x=1,y=0,label='Average in highest quartile',hjust=1) +
  scale_x_continuous(limits=c(0,1.5)) +
  theme_classic() +
  theme(legend.position='none',
        axis.ticks=element_blank(),
        axis.line=element_blank(),
        axis.text.x=element_blank(),
        axis.title=element_blank(),
        text=element_text(size=20))
```

The `diso` variables all increase dramatically, while `aoj11` shows a dramatic decrease. This suggests that high values of the index indicate not being particularly concerned about crime and violence; we'll have to reverse the scale.

```{r}
all_pc1$norm <- scale(-all_pc1$avg) # sign flip -- now high values mean more fear
predict_data <- data.frame(fear10 = c(0,1), vic1ext = c(2,1),
                           vic1exta=c(0,max(my_data$vic1exta)),
                           vic1hogar=c(2,1),aoj11=c(1,4),pese1=c(3,1),
                           pese2=c(3,1),aoj17=c(4,1),diso7=c(5,1),diso8=c(5,1),
                           diso10=c(5,1),diso18=c(5,1),diso14=c(5,1),
                           diso16=c(5,1),diso17=c(5,1),vicbar1=c(2,1),
                           vicbar1f=c(3,1),vicbar3=c(2,1),vicbar4=c(2,1),
                           vicbar7=c(2,1),a4=c(0,1))

minmax <- rowMeans(sapply(1:5, function(i) predict(pr[[i]],predict_data)[,1]))
minmax <- (mean(all_pc1$avg) - minmax) / sd(all_pc1$avg)
# quantile(all_pc1$norm)
fear <- data.frame(w=all_pc1$norm) 
```

```{r}
predict_data2 <- data.frame(diag(21))
names(predict_data2) <- names(predict_data)
# right now, you have something that would feed 0's and 1's into your score
# prediction -- this doesn't really make sense because many of your questions
# are on a scale from 1-2 or 1-4. Here's one (slightly clunky) way to fix it:
predict_data2$fear10 <- 4 - 3*predict_data2$fear10 
predict_data2$vic1ext <- 2 - predict_data2$vic1ext
predict_data2$vic1exta <- 20*predict_data2$vic1exta
predict_data2$vic1hogar <- 2 - predict_data2$vic1hogar
predict_data2$aoj11 <- 1 + 3*predict_data2$aoj11
predict_data2$aoj17 <- 4 - 3*predict_data2$aoj17
predict_data2$pese1 <- 3 - 2*predict_data2$pese1
predict_data2$pese2 <- 3 - 2*predict_data2$pese2
predict_data2$diso7 <- 5 - 4*predict_data2$diso7
predict_data2$diso8 <- 5 - 4*predict_data2$diso8
predict_data2$diso10 <- 5 - 4*predict_data2$diso10
predict_data2$diso18 <- 5 - 4*predict_data2$diso18
predict_data2$diso14 <- 5 - 4*predict_data2$diso14
predict_data2$diso16 <- 5 - 4*predict_data2$diso16
predict_data2$diso17 <- 5 - 4*predict_data2$diso17
predict_data2$vicbar1 <- 2 - predict_data2$vicbar1
predict_data2$vicbar1f <- 3 - 2*predict_data2$vicbar1f
predict_data2$vicbar3 <- 2 - predict_data2$vicbar3
predict_data2$vicbar4 <- 2 - predict_data2$vicbar4
predict_data2$vicbar7 <- 2 - predict_data2$vicbar7
```

Now, in each row of predict_data2, all values are those that will give a lower score, except for one.

```{r}
scores <- rowMeans(sapply(1:5, function(i) predict(pr[[i]],predict_data2)[,1]))
scores <- (mean(all_pc1$avg)-scores)/sd(all_pc1$avg)
```

Now you can see that all of the scores are reasonably close to the minimum value stored in minmax.
```{r}
diff <- data.frame(d=(scores-minmax[1]), n = names(predict_data))
ggplot(diff, aes(x=n,y=d)) + 
  geom_bar(stat = 'identity', fill = 'skyblue') + 
  coord_flip() + 
  theme_classic() + 
  theme(text = element_text(size=20), 
        axis.title.y = element_blank(),
        axis.title.x = element_blank())
```

All variables seem to have an effect, with a somewhat smaller one for `fear10` (avoided walking through dangerous areas), and larger ones for the `diso` variables and `vic1exta` (frequency of victimization)

```{r,warning=FALSE,message=FALSE}
fear10 <- data.frame(q=my_data$fear10,w=fear$w)
fear10 <- na.omit(fear10)
my_lm <- lm(data=fear10, q ~ w)
ggplot(fear10,aes(x=w,y=q)) +
  geom_jitter(color='tomato3',size=5,alpha=0.2) +
  theme_classic() +
  geom_smooth(method='lm',size=2,color='royalblue') +
  annotate('text',label=paste("R-sq:",round(summary(my_lm)$r.squared,2)),  
  size=10,x=-1.5,y=1,hjust=0,vjust=0,color='royalblue') +
  ylab('fear10') +
  xlab('Fear index') +
  theme(text=element_text(size=20))
```

We can make this all more rigorous by examining p-values:

```{r}
lmp <- function (modelobject) {
  if (class(modelobject) != "lm") stop("Not an object of class 'lm' ")
    f <- summary(modelobject)$fstatistic
    p <- pf(f[1],f[2],f[3],lower.tail=F)
    attributes(p) <- NULL
    return(p)
}
lm_data <- my_data[,c('vic1exta','aoj11','aoj17','pese1','pese2',
                      'diso7','diso8','diso10','diso14','diso16','diso17',
                      'diso18','vicbar3')]
sapply(lm_data, function(x) lmp(lm(x ~ all_pc1$norm)))

bin_cor <- function(var) {
  tt <- t.test(all_pc1$norm[var==min(var,na.rm=TRUE)],
               all_pc1$norm[var==max(var,na.rm=TRUE)])
  tt$p.value
}

bin_vars <- c('fear10','vic1ext','vic1hogar','vicbar1','vicbar3',
              'vicbar4','vicbar7')
sapply(bin_vars,function(x) bin_cor(my_data[,x]))
```

Finally, check these values against the comments in `make_indices.R` to be sure we're calculating the same index:

```{r}
quantile(all_pc1$norm)
```


```{r,message=FALSE}
set.seed(12345) # for reproducibility

incl <- c('vic44','fear10','vic1ext',
         'vic1exta','vic1hogar','aoj11','pese1','pese2','aoj17','diso7',
         'diso8','diso10','diso18','diso14','diso16','diso17','vicbar1', 
         'vicbar1f','vicbar3','vicbar4','vicbar7')

hnd_data <- lapop.2014.HND[,incl]
slv_data <- lapop.2014.SLV[,incl]
gtm_data <- lapop.2014.GTM[,incl]
my_data <- rbind(gtm_data,hnd_data,slv_data)

my_data$vic1exta[my_data$vic1exta == 999999] <- 0
is.na(my_data[my_data>30]) <- TRUE

my_imp <- mice(my_data, printFlag = F)
pr <- lapply(1:5,function(x) prcomp(complete(my_imp,x),scale=FALSE,center=TRUE))

all_pc1 <- data.frame(llply(1:5, function(i) pr[[i]]$x[,1]))
names(all_pc1) <- c('imp1', 'imp2', 'imp3', 'imp4', 'imp5')
all_pc1$imp1 <- -all_pc1$imp1
all_pc1$imp4 <- -all_pc1$imp4
all_pc1$avg <- rowMeans(all_pc1)
all_pc1$norm <- scale(-all_pc1$avg) # sign flip -- now high values mean more fear
```

Let's see how people's level of fear correlates with their answers to `a4` -- the most serious problem facing the country. This is a little tricky, since the coding is consistent between countries (e.g. `a4=4` means poverty everywhere), but not all of the possible answers were on the questionnaire for any given country. If any country-specific answers turn out to be significant, we'll have to make sure they're significant at a country level as well.

```{r}
fear <- data.frame(w=all_pc1$norm)
a4 <- c(lapop.2014.HND$a4,lapop.2014.SLV$a4,lapop.2014.GTM$a4)
a4[a4==888888] <- NA
a4[a4==988888] <- NA
fear$a4 <- a4

categ_unsorted <- function(f,x,categ) {
  # For a data from f with a continuous variable 'x' and an unsorted 
  # categorical variable 'categ', test whether the value of x is 
  # significantly higher or lower for each value of categ than for the
  # rest of the population. Do this using a two-sample t-test.
  result <- data.frame(var=character(),val=integer(),pval=double(),
                       mean=double(),othermean=double())
  for(q in unique(na.omit(f[,categ]))) {
    if(sum(f[,categ]==q,na.rm=TRUE) > 1) {
      yes <- fear[fear[,categ]==q,x]
      no <- fear[fear[,categ]!=q,x]
      tt <- t.test(yes,no,na.rm=TRUE)
      newrow <- data.frame(var=categ,val=q,pval=tt$p.value,
                       mean=tt$estimate[[1]],
                       othermean=tt$estimate[[2]])
      result <- rbind(result,newrow)
    }
  }
  result
}

categ_unsorted(fear,'w','a4')
```

So it looks like the only strong correlation here is that people who are concerned about human rights violations are significantly *less* fearful than the rest of the population. 

Another unordered categorical variable would be `for1n` (Country with the most influence in the region).

```{r}
for1n <- c(lapop.2014.HND$for1n,lapop.2014.SLV$for1n,lapop.2014.GTM$for1n)
for1n[for1n==888888] <- NA
for1n[for1n==988888] <- NA
fear$for1n <- for1n
categ_unsorted(fear,'w','for1n')
mean(for1n==4,na.rm=TRUE)
mean(for1n==2,na.rm=TRUE)
```

So the 71% of people who feel that the USA (`for1n=4`) is the most influential country in the region are significantly more fearful than the remaining 29%, while the 9% who think Japan has the most influence (`for1n==2`) are less fearful than the remaining 91%. Everything else turns out not to be significant. I don't really know what makes the people who perceive a high Japanese influence different from everyone else, but it's a pretty big difference.

