---
title: "Comparing survey contents between different waves"
output:
  rmarkdown::html_vignette:
    toc: yes
    fig_width: 8
    fig_height: 6
vignette: >
  %\VignetteIndexEntry{Comparing survey contents between different waves}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, 
                      message = FALSE, 
                      warning = FALSE, 
                      collapse = FALSE,
                      comment = "#>",
                      fig.align = "center")
#knitr::opts_chunk$set(fig.width = 8, fig.height = 6)
set.seed(1)
extrafont::loadfonts(quiet=TRUE)
options(scipen = 999) # turn-off scientific notation like 1e+48


```


```{r,message=FALSE}
library(unhcRstyle)
library(ggplot2)
library(mice)
library(plyr) 
library(GGally)
library(ggfortify)
library(stargazer)
library(lubridate)
library(VennDiagram)
library(RColorBrewer)
library(AmericasBarometer)
#source('../utils/make_trends.R')

mainDir <- getwd()
## If you save your analysis under vignette folder...
mainDirroot <- substring(mainDir, 0 , nchar(mainDir) - 10)


lapop.2014.GTM <- read.csv(paste0(mainDirroot, "/data-raw/lapop.2014.GTM.csv"))
lapop.2014.SLV <- read.csv(paste0(mainDirroot, "/data-raw/lapop.2014.SLV.csv"))
lapop.2014.HND <- read.csv(paste0(mainDirroot, "/data-raw/lapop.2014.HND.csv"))

# Get useful non-numeric variables
date_time <- rbind(lapop.2014.GTM[,c('pais','fecha' #,'ti'
                                     )],
                   lapop.2014.SLV[,c('pais','fecha'#,'ti'
                                     )],
                   lapop.2014.HND[,c('pais','fecha'#,'ti'
                                     )])

```




```{r,message=FALSE}
fixdata <- function(x) {
  y <- x
  # vic1exta asks about the frequency of crime victimization and is NR if 
  # respondant was never victimized; change this to zero
  y$vic1exta[y$vic1exta == 999999] <- 0
  # vicbar1f is similar; asks about the number of occurrences of burglaries
  y$vicbar1f[y$vicbar1f == 999999] <-  0
  # vicbar7f asks about the number of occurences of murders (SLV only)
  if ('vicbar7f' %in% names(y)) {
    y$vicbar7f[y$vicbar7f == 999999] <- 0
  }
  # set uniq_id to 0 (so it doesn't become NA)
  y$uniq_id <- 0
  # somehow ages got imported as strings
  y$q2 <- as.numeric(y$q2)
  # set non-responses to NA
  y[y > 800000] <- NA
  y
}

lapop.2014.GTM <- fixdata(lapop.2014.GTM)
lapop.2014.SLV <- fixdata(lapop.2014.SLV)
lapop.2014.HND <- fixdata(lapop.2014.HND)

# create a combined dataset
common <- Reduce(intersect,list(names(lapop.2014.GTM),names(lapop.2014.SLV),
                                names(lapop.2014.HND)))
lapop.2014.all <- rbind(lapop.2014.GTM[,common],lapop.2014.SLV[,common],
                        lapop.2014.HND[,common])

```

# Basic features 

```{r,message=FALSE,warning=FALSE}
basic <- data.frame(country=c('Guatemala','El Salvador','Honduras'),
                    nq=c(ncol(lapop.2014.GTM),ncol(lapop.2014.SLV),
                         ncol(lapop.2014.HND)),
                    nr=c(nrow(lapop.2014.GTM),nrow(lapop.2014.SLV),
                         nrow(lapop.2014.HND)))
# num of questions
ggplot(basic,aes(x=reorder(country, nq),
                 y=nq #, fill=country
                 )) +
  geom_bar(stat="identity", fill = unhcr_blue) +
  ggtitle("Number of questions") +
  geom_text(aes(y=nq,label=nq,vjust=1.5),size=6) +
 # scale_fill_manual(values=brewer.pal(3,"Dark2")) +
  coord_flip() +
  labs(title = "Number of available analysis variables per country & Wave", 
            # subtitle = " ",
             x = " ", 
             y = "",
             caption = "Latin American Public Opinion Project / Vanderbilt University") +
          unhcRstyle::unhcr_theme() +
          theme(axis.text = element_text(size = 6),
                legend.position = "none",
                panel.grid.major.x = element_line(color = "#cbcbcb"), 
                panel.grid.major.y = element_blank()) +### changing grid line that should appear) +
        geom_hline(yintercept = 0, size = 0.7, colour = "#333333")
```




```{r,message=FALSE}
# num of responses
ggplot(basic,aes(x=reorder(country, nq),
                 y=nr #,fill=country
                 )) +
  geom_bar(stat="identity", fill = unhcr_blue) +
  geom_text(aes(y=nr,label=nr,vjust=1.5),size=6) +
  #  scale_fill_manual(values=brewer.pal(3,"Dark2")) +
  coord_flip() +
  labs(title = "Number of respondents per country & Wave", 
            # subtitle = " ",
             x = " ", 
             y = "",
             caption = "Latin American Public Opinion Project / Vanderbilt University") +
          unhcRstyle::unhcr_theme() +
          theme(axis.text = element_text(size = 6),
                legend.position = "none",
                panel.grid.major.x = element_line(color = "#cbcbcb"), 
                panel.grid.major.y = element_blank()) +### changing grid line that should appear) +
        geom_hline(yintercept = 0, size = 0.7, colour = "#333333")
```




```{r,message=FALSE, warning=FALSE}
# Venn diagram of overlapping questions
#plot.new()
venn.plot <- draw.triple.venn(
                 ncol(lapop.2014.GTM),ncol(lapop.2014.SLV),ncol(lapop.2014.GTM),
                 sum(names(lapop.2014.GTM) %in% names(lapop.2014.SLV)),
                 sum(names(lapop.2014.SLV) %in% names(lapop.2014.HND)),
                 sum(names(lapop.2014.GTM) %in% names(lapop.2014.HND)),
                 ncol(lapop.2014.all),
                 category=c('Guatemala','El Salvador','Honduras'),
                 fill=brewer.pal(3,"Dark2")[c(2,1,3)],
                 ## Font size & type
                 cex=rep(3,7),
                 cat.cex=rep(3,3),
                 cat.fontfamily="Lato",
                 
                 alpha=rep(0.1,3),
                 col=rep(NA,3),
                 cat.just=list(c(0,0.5),
                               c(1,0.5),
                               c(0.5,0)),
                 euler.d=TRUE,
                 scaled=TRUE,
                	main = "Overlapping Questions",
                	sub = "Between countries & Wave",
                	main.cex = 2,
                	sub.cex = 1)

grid.draw(venn.plot);
grid.newpage();

```




```{r,message=FALSE}
# Questions with time series

# # Create test data.
# ts = data.frame(count=c(ncol(lapop.trends), ncol(lapop.2014.all)-ncol(lapop.trends)), 
#                 category=c('trend','notrend'))
# 
# # Add addition columns, needed for drawing with geom_rect.
# ts$fraction = ts$count / sum(ts$count)
# ts = ts[order(ts$fraction), ]
# ts$ymax = cumsum(ts$fraction)
# ts$ymin = c(0, head(ts$ymax, n=-1))
# ggplot(ts, aes(fill=category, ymax=ymax, ymin=ymin, xmax=4, xmin=2)) +
#      geom_rect() +
#      coord_polar(theta="y") +
#      xlim(c(0, 4)) +
#      annotate('text',x=4,y=0.25,
#               label='2004-2014 time series',size=10,hjust=0.85) +
#      annotate('text',x=4,y=0.82,
#               label='No time series',size=10,hjust=0.3) +
#      theme_classic() +
#      theme(legend.position='none',
#            axis.ticks=element_blank(),
#            axis.text=element_blank(),
#            axis.line=element_blank(),
#            axis.title=element_blank())
```

# Missing values 

```{r message=FALSE, warning=FALSE}
# Average missing values
mean(is.na(lapop.2014.GTM)) # 8.9% in GTM
mean(is.na(lapop.2014.SLV)) # 9.0% in SLV
mean(is.na(lapop.2014.HND)) # 9.9% in HND
mean(is.na(lapop.2014.all)) # 8.2% in overlap
# Missing values per row
missing_row <- rbind(data.frame(x=rowMeans(is.na(lapop.2014.GTM)),country='GTM'),
                     data.frame(x=rowMeans(is.na(lapop.2014.SLV)),country='SLV'),
                     data.frame(x=rowMeans(is.na(lapop.2014.HND)),country='HND'),
                     data.frame(x=rowMeans(is.na(lapop.2014.all)),country='all'))
ggplot(missing_row,aes(x=x,fill=country)) +
  geom_density(alpha=0.2) +
  scale_fill_manual(values=brewer.pal(4,"Dark2")[c(2,1,3,4)]) +
  scale_x_continuous(limits=c(0,0.3)) +
  labs(title = "Missing values per row", 
            # subtitle = " ",
             x = " ", 
             y = "",
             caption = "Latin American Public Opinion Project / Vanderbilt University") +
          unhcRstyle::unhcr_theme() +
          theme(axis.text = element_text(size = 6),
                #legend.position = "none",
                panel.grid.major.x = element_line(color = "#cbcbcb"), 
                panel.grid.major.y = element_blank()) +### changing grid line that should appear) +
        geom_hline(yintercept = 0, size = 0.7, colour = "#333333")

```


```{r message=FALSE, warning=FALSE}
# head(sort(colMeans(is.na(lapop.2014.GTM)),decreasing=TRUE))
# head(sort(colMeans(is.na(lapop.2014.SLV)),decreasing=TRUE))
# head(sort(colMeans(is.na(lapop.2014.HND)),decreasing=TRUE))
head(sort(colMeans(is.na(lapop.2014.all)),decreasing=TRUE),10)
# quantile(colMeans(is.na(lapop.2014.GTM)))
# quantile(colMeans(is.na(lapop.2014.SLV)))
# quantile(colMeans(is.na(lapop.2014.HND)))
# quantile(colMeans(is.na(lapop.2014.all)))
```

The most common missing responses are for:

- `vb101`: Reason for nullifying ballot (98.3%)
- `exc14`: Paid a bribe at the courts (95.6%)
- `vic2`: Type of victimization (82.3%)
- `vic2aa`: Location of victimization (82.2%)
- `exc11`: Bribe required to process document in municipal government (77.5%)
- `vb4new`: Reason for not voting (74.7%)

In general, questions with many missing answers were requesting information about fairly unusual experiences. In all cases, 50% of questions have missing-answer rates of less than 1-2%.

# Survey dates 

For some reason, a single response in Guatemala was from 1/1/2012, while the rest are from the spring of 2014

```{r message=FALSE, warning=FALSE}
Sys.setlocale("LC_TIME", "usa")
date_time <- lapop.2014.all[-1407,] # remove outlier
date_time$date <- readr::parse_date(as.character(date_time$fecha))
countries <- c(NA,'Guatemala','El Salvador','Honduras')

date_time$country <- countries[date_time$pais]


ggplot(date_time,
       aes(x=date,
           fill=country)) +
  geom_density(alpha=0.2,
               adjust=0.25) +
  
  scale_fill_manual(values=brewer.pal(3,"Dark2")) +
  labs(title = "Survey data collection date", 
            # subtitle = " ",
             x = " ", 
             y = "",
             caption = "Latin American Public Opinion Project / Vanderbilt University") +
          unhcRstyle::unhcr_theme() +
          theme(axis.text = element_text(size = 6),
                #legend.position = "none",
                panel.grid.major.x = element_line(color = "#cbcbcb"), 
                panel.grid.major.y = element_blank()) +### changing grid line that should appear) +
        geom_hline(yintercept = 0, size = 0.7, colour = "#333333")
```




```{r,message=FALSE}

# ggplot(date_time,aes(x=ti,fill=country)) +
#   geom_density(alpha=0.2,adjust=0.25) +
#   scale_x_continuous(limits=c(15,90)) +
#   scale_fill_manual(values=brewer.pal(3,"Dark2")) +
#   theme_classic() +
#   xlab('Survey time (minutes)') +
#   theme(axis.ticks=element_blank(),
#         axis.text.y=element_blank(),
#         axis.line=element_blank(),
#         axis.title.y=element_blank(),
#         text=element_text(size=20))
```

Most people took 30-60 minutes to complete the survey.

# Basic Demographics 

```{r message=FALSE, warning=FALSE}
#mean(lapop.2014.all$sex==2) # 51.8% female
ages <- lapop.2014.all[,c('pais','q2')]
ages$country <- countries[ages$pais]


ggplot(ages,aes(x=q2,fill=country)) +
  geom_density(alpha=0.2) +
  scale_fill_manual(values=brewer.pal(3,"Dark2")) +
  labs(title = "Breakdown of main respondent age", 
            # subtitle = " ",
             x = "Age (years) ", 
             y = "",
             caption = "Latin American Public Opinion Project / Vanderbilt University") +
          unhcRstyle::unhcr_theme() +
          theme(axis.text = element_text(size = 6),
               # legend.position = "none",
                panel.grid.major.x = element_line(color = "#cbcbcb"), 
                panel.grid.major.y = element_blank()) +### changing grid line that should appear) +
        geom_hline(yintercept = 0, size = 0.7, colour = "#333333")

```