---
title: "Trends analysis"
output:
  rmarkdown::html_vignette:
    toc: yes
    fig_width: 8
    fig_height: 6
vignette: >
  %\VignetteIndexEntry{Trends analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, 
                      message = FALSE, 
                      warning = FALSE, 
                      collapse = FALSE,
                      comment = "#>",
                      fig.align = "center")
```

```{r}
library(AmericasBarometer)
library(plyr)
library(ggplot2)
library(stringr)
library(RColorBrewer)

mainDir <- getwd()
## If you save your analysis under vignette folder...
mainDirroot <- substring(mainDir, 0 , nchar(mainDir) - 10)


lapop.trends <- read.csv(paste0(mainDirroot, "/data/dataLAPOP.csv"))

#table(lapop.trends$year, useNA= "ifany") 
## let's put 2019 & 2018 together

lapop.trends$year[lapop.trends$year=="2019"] <- "2018"
lapop.trends$year <- as.integer(lapop.trends$year)
lapop.trends$wave <- paste0(lapop.trends$ctrycollect, lapop.trends$year)

```



Time trends are here assemble by stitching together individual files for the three Northern Triangle countries.


*NOTE:* While indicator names might be consistent from year to year, values aren't necessarily. In 2014, `cp5` (tried to solve a community problem) was on a scale of 1-4 (4=never). Earlier years just used 1=yes, 2=no, so it looks like there was a big jump in 2008.

## Migration	

###  q14 Intends to Live or Work Abroad
	
```{r}
#table(lapop.trends$q14, useNA= "ifany") 
q14  <- ddply(lapop.trends,~year,summarize,x=sum(!is.na(q14 ) & q14 ==1),
            n=sum(!is.na(q14) & (q14==2 | q14==1)))

AmericasBarometer::trend_plot(q14,
                              'Intends to Live or Work Abroad',
                              '  ')
```	


	
	


```{r}
###  q10a	Receives Remittances
#Problem with variable encoding in 2006...
#table(lapop.trends$q10a, useNA= "ifany")
# table(lapop.trends$q10a, lapop.trends$wave, useNA= "ifany") 
# q10  <- ddply(lapop.trends,~year,summarize,x=sum(!is.na(q10 ) & q10 ==1),
#             n=sum(!is.na(q10) & (q10==2 | q10==1)))
# 
# AmericasBarometer::trend_plot(q10,
#                               'Receives Remittances',
#                               '  ')
```	

## Civic and Political Engagement	

## Political ideology
The distribution of political ideology (`l1`) is interesting; the largest group puts themselves smack in the middle, rather than tilting left or right. This centrist group (most likely politically disengaged) has stayed pretty constant over the last decade, while on average the right is losing out to the left.
```{r}
#bar_plot(data.frame(x=na.omit(lapop.trends$l1)))
#table(lapop.trends$l1, lapop.trends$wave, useNA= "ifany")
# Three logical groupings: left, center, right
l1 <- ddply(lapop.trends,~year,summarize,
            left=sum(l1<5,na.rm=TRUE) / sum(l1<11,na.rm=TRUE),
            center=sum(l1==5 | l1==6,na.rm=TRUE) / sum(l1<11,na.rm=TRUE),
            right=sum(l1>6,na.rm=TRUE) / sum(l1<11,na.rm=TRUE))

AmericasBarometer::multi_lines(l1,
                              'Political ideology',
                              '  ')
```

###  cp2	Requested Help from a Member of the Legislature
	
```{r}
#table(lapop.trends$cp2, useNA= "ifany") 
cp2  <- ddply(lapop.trends,~year,summarize,x=sum(!is.na(cp2 ) & cp2 ==1),
            n=sum(!is.na(cp2) & (cp2==2 | cp2==1)))

AmericasBarometer::trend_plot(cp2,
                              'Requested Help from a Member of the Legislature',
                              '  ')
```	

###  cp13	Attendance at Meetings of Political Parties
	
```{r}
#table(lapop.trends$cp13, useNA= "ifany") 
cp13 <- ddply(lapop.trends,~year,summarize,
             weekly=sum(cp13==1,na.rm=TRUE) / sum(cp13 < 5,na.rm=TRUE),
             monthly=sum(cp13==2,na.rm=TRUE) / sum(cp13 < 5,na.rm=TRUE),
             annual=sum(cp13==3,na.rm=TRUE) / sum(cp13 < 5,na.rm=TRUE),
             never=sum(cp13==4,na.rm=TRUE) / sum(cp13 < 5,na.rm=TRUE))

AmericasBarometer::multi_lines(cp13,
                              'Attendance at Meetings of Political Parties',
                              '  ')  
```	



### cp20	Attendance at Meetings of Women's Group
	
```{r}
#table(lapop.trends$cp20, useNA= "ifany") 
cp20 <- ddply(lapop.trends,~year,summarize,
             weekly=sum(cp20==1,na.rm=TRUE) / sum(cp20 < 5,na.rm=TRUE),
             monthly=sum(cp20==2,na.rm=TRUE) / sum(cp20 < 5,na.rm=TRUE),
             annual=sum(cp20==3,na.rm=TRUE) / sum(cp20 < 5,na.rm=TRUE),
             never=sum(cp20==4,na.rm=TRUE) / sum(cp20 < 5,na.rm=TRUE))

AmericasBarometer::multi_lines(cp20,
                              'Attendance at Meetings of Women  Group',
                              '  ') 
```	

### cp6	Attendance at Meetings of Religious Organization


The number of occasional churchgoers has been pretty constant (and low); it looks like the population of weekly churchgoers has been growing while non-attenders have been dropping since 2008. What changed in 2008?

```{r}
#table(lapop.trends$cp6, useNA= "ifany") 

cp6 <- ddply(lapop.trends,~year,summarize,
             weekly=sum(cp6==1,na.rm=TRUE) / sum(cp6 < 5,na.rm=TRUE),
             monthly=sum(cp6==2,na.rm=TRUE) / sum(cp6 < 5,na.rm=TRUE),
             annual=sum(cp6==3,na.rm=TRUE) / sum(cp6 < 5,na.rm=TRUE),
             never=sum(cp6==4,na.rm=TRUE) / sum(cp6 < 5,na.rm=TRUE))

AmericasBarometer::multi_lines(cp6,
                              'Attendance at Meetings of Religious Organization',
                              '  ')
```

### cp7	Attendance at Meetings of Parent Association
	
```{r}
#table(lapop.trends$cp7, useNA= "ifany") 
cp7 <- ddply(lapop.trends,~year,summarize,
             weekly=sum(cp7==1,na.rm=TRUE) / sum(cp7 < 5,na.rm=TRUE),
             monthly=sum(cp7==2,na.rm=TRUE) / sum(cp7 < 5,na.rm=TRUE),
             annual=sum(cp7==3,na.rm=TRUE) / sum(cp7 < 5,na.rm=TRUE),
             never=sum(cp7==4,na.rm=TRUE) / sum(cp7 < 5,na.rm=TRUE))

AmericasBarometer::multi_lines(cp7,
                              'Attendance at Meetings of Parent Association',
                              '  ')
```	

### cp8	Attendance at Meetings for Community Improvements
	
```{r}
#table(lapop.trends$cp8, useNA= "ifany") 
cp8 <- ddply(lapop.trends,~year,summarize,
             weekly=sum(cp8==1,na.rm=TRUE) / sum(cp8 < 5,na.rm=TRUE),
             monthly=sum(cp8==2,na.rm=TRUE) / sum(cp8 < 5,na.rm=TRUE),
             annual=sum(cp8==3,na.rm=TRUE) / sum(cp8 < 5,na.rm=TRUE),
             never=sum(cp8==4,na.rm=TRUE) / sum(cp8 < 5,na.rm=TRUE))

AmericasBarometer::multi_lines(cp8,
                              'Attendance at Meetings for Community Improvements',
                              '  ')
```	

### prot3	Participated in a Protest
	
```{r}
#table(lapop.trends$prot3, useNA= "ifany") 

# table(lapop.trends$prot3, lapop.trends$wave, useNA= "ifany") 
prot3  <- ddply(lapop.trends,~year,summarize,x=sum(!is.na(prot3 ) & prot3 ==1),
            n=sum(!is.na(prot3) & (prot3==2 | prot3==1)))

AmericasBarometer::trend_plot(prot3,
                              'Participated in a Protest',
                              '  ')
```	

### vb10	Identifies with a Political Party
	
```{r}
#table(lapop.trends$vb10, useNA= "ifany") 
vb10  <- ddply(lapop.trends,~year,summarize,x=sum(!is.na(vb10 ) & vb10 ==1),
            n=sum(!is.na(vb10) & (vb10==2 | vb10==1)))

AmericasBarometer::trend_plot(vb10,
                              'Identifies with a Political Party',
                              '  ')
```	

### vb2	Voted in Last Presidential Election
	
```{r}
#table(lapop.trends$vb2, useNA= "ifany") 
vb2  <- ddply(lapop.trends,~year,summarize,x=sum(!is.na(vb2 ) & vb2 ==1),
            n=sum(!is.na(vb2) & (vb2==2 | vb2==1)))

AmericasBarometer::trend_plot(vb2,
                              'Declared that voted in Last Presidential Election',
                              '  ')
```	

### vb20	Vote Intention for President
	
```{r}
#table(lapop.trends$vb20, useNA= "ifany") 

# 1             No votaría
# 2             Votaría por el candidato o partido del actual presidente/administración
#3             Votaría por algún candidato o partido diferente                 del actual gobierno
#4             Iría a votar pero dejaría la boleta en blanco o                 la anularía

vb20 <- ddply(lapop.trends,~year,summarize,
             NoVote=sum(vb20==1,na.rm=TRUE) / sum(vb20 < 5,na.rm=TRUE),
             ActualAdministration=sum(vb20==2,na.rm=TRUE) / sum(vb20 < 5,na.rm=TRUE),
             OtherAdministration=sum(vb20==3,na.rm=TRUE) / sum(vb20 < 5,na.rm=TRUE),
             Blank=sum(vb20==4,na.rm=TRUE) / sum(vb20 < 5,na.rm=TRUE))

AmericasBarometer::multi_lines(vb20,
                              'Vote Intention for President',
                              '  ')
```	


The question about approval of political demonstrations (`e5`) lets people express a range from strong approval to strong disapproval -- most are somewhere in the middle. People who are actively approving of political demonstrations seem to be  losing ground to those who are either wishy-washy or disapproving

Again, 2008 looks like some kind of turning point.
```{r}
# This reminds me a bit of the political ideology chart -- lots concentrated at either end
e5 <- ddply(lapop.trends,~year,summarize,
            disapprove=sum(e5<4,na.rm=TRUE) / sum(e5<11,na.rm=TRUE),
            meutral=sum(e5>3 & e5<6,na.rm=TRUE) / sum(e5<11,na.rm=TRUE),
            approve=sum(e5>7,na.rm=TRUE) / sum(e5<11,na.rm=TRUE))

AmericasBarometer::multi_lines(e5,
                              'Approval of Those Participating in Legal Demonstration',
                              '  ')
```


	
## Climate Change/Disaster Risk Reduction	

### drk1	Likelihood of Death or Harm from Natural Disaster
	
```{r}
#table(lapop.trends$drk1, useNA= "ifany") 
drk1 <- ddply(lapop.trends,~year,summarize,
             Not_probable=sum(drk1==1,na.rm=TRUE) / sum(drk1 < 5,na.rm=TRUE),
             Unlikely=sum(drk1==2,na.rm=TRUE) / sum(drk1 < 5,na.rm=TRUE),
             Possible=sum(drk1==3,na.rm=TRUE) / sum(drk1 < 5,na.rm=TRUE),
             Very_Possible=sum(drk1==4,na.rm=TRUE) / sum(drk1 < 5,na.rm=TRUE))

AmericasBarometer::multi_lines(drk1,
                              'Likelihood of Death or Harm from Natural Disaster',
                              '  ')
```	

### dst1b	Spend More to Enforce Building Codes
	
```{r}
#table(lapop.trends$dst1b, useNA= "ifany") 
dst1b <- ddply(lapop.trends,~year,summarize,
            disapprove=sum(dst1b<4,na.rm=TRUE) / sum(dst1b<11,na.rm=TRUE),
            meutral=sum(dst1b>3 & dst1b<6,na.rm=TRUE) / sum(dst1b<11,na.rm=TRUE),
            approve=sum(dst1b>7,na.rm=TRUE) / sum(dst1b<11,na.rm=TRUE))

AmericasBarometer::multi_lines(dst1b,
                              'Approval to Spend More to Enforce Building Codes',
                              '  ')
```	

	
```{r}

### env1c	Higher Priority for Protecting Environment or Promoting Economic Growth
#table(lapop.trends$env1c, useNA= "ifany") 
# env1c <- ddply(lapop.trends,~year,summarize,
#             disapprove=sum(env1c<4,na.rm=TRUE) / sum(env1c<8,na.rm=TRUE),
#             meutral=sum(env1c>3 & env1c<6,na.rm=TRUE) / sum(env1c<8,na.rm=TRUE),
#             approve=sum(env1c>7,na.rm=TRUE) / sum(env1c<8,na.rm=TRUE))
# 
# AmericasBarometer::multi_lines(env1c,
#                               'Higher Priority for Protecting Environment or Promoting Economic Growth',
#                               '  ')
```	

### env2b	Seriousness of Climate Change
	
```{r}
#table(lapop.trends$env2b,lapop.trends$wave, useNA= "ifany") 
env2b <- ddply(lapop.trends,~year,summarize,
             Very_serious=sum(env2b==1,na.rm=TRUE) / sum(env2b < 5,na.rm=TRUE),
             Serious=sum(env2b==2,na.rm=TRUE) / sum(env2b < 5,na.rm=TRUE),
             Little_serious=sum(env2b==3,na.rm=TRUE) / sum(env2b < 5,na.rm=TRUE),
             Not_Serious=sum(env2b==4,na.rm=TRUE) / sum(env2b < 5,na.rm=TRUE))

AmericasBarometer::multi_lines(env2b,
                              'Seriousness of Climate Change',
                              '  ')
```	


	
## Corruption	

### exc11	Asked to Pay Bribe to Process Document in Municipality
	
```{r}
#table(lapop.trends$exc11, useNA= "ifany") 
exc11  <- ddply(lapop.trends,~year,summarize,x=sum(!is.na(exc11 ) & exc11 ==1),
            n=sum(!is.na(exc11) & (exc11==0 | exc11==1)))

AmericasBarometer::trend_plot(exc11,
                              'Asked to Pay Bribe to Process Document in Municipality',
                              '  ')
```	

### exc13	Asked to Pay a Bribe at Work
	
```{r}
#table(lapop.trends$exc13, useNA= "ifany") 
exc13  <- ddply(lapop.trends,~year,summarize,x=sum(!is.na(exc13 ) & exc13 ==1),
            n=sum(!is.na(exc13) & (exc13==0 | exc13==1)))

AmericasBarometer::trend_plot(exc13,
                              'Asked to Pay a Bribe at Work',
                              '  ')
```	

### exc14	Asked to Pay a Bribe to the Courts
	
```{r}
#table(lapop.trends$exc14, useNA= "ifany") 
exc14  <- ddply(lapop.trends,~year,summarize,x=sum(!is.na(exc14 ) & exc14 ==1),
            n=sum(!is.na(exc14) & (exc14==0 | exc14==1)))

AmericasBarometer::trend_plot(exc14,
                              'Asked to Pay a Bribe to the Courts',
                              '  ')
```	

### exc15	Asked to Pay a Bribe to use Public Health Services
	
```{r}
#table(lapop.trends$exc15, useNA= "ifany") 
exc15  <- ddply(lapop.trends,~year,summarize,x=sum(!is.na(exc15 ) & exc15 ==1),
            n=sum(!is.na(exc15) & (exc15==0 | exc15==1)))

AmericasBarometer::trend_plot(exc15,
                              'Asked to Pay a Bribe to use Public Health Services',
                              '  ')
```	

### exc16	Asked to Pay a Bribe at School
	
```{r}
#table(lapop.trends$exc16, useNA= "ifany") 
exc16  <- ddply(lapop.trends,~year,summarize,x=sum(!is.na(exc16 ) & exc16 ==1),
            n=sum(!is.na(exc16) & (exc16==0 | exc16==1)))

AmericasBarometer::trend_plot(exc16,
                              'Asked to Pay a Bribe at School',
                              '  ')
```	

### exc18	Paying a Bribe is Justified
	
```{r}
#table(lapop.trends$exc18, useNA= "ifany") 
exc18  <- ddply(lapop.trends,~year,summarize,x=sum(!is.na(exc18 ) & exc18 ==1),
            n=sum(!is.na(exc18) & (exc18==0 | exc18==1)))

AmericasBarometer::trend_plot(exc18,
                              'Paying a Bribe is Justified',
                              '  ')
```	

### exc2	Police Officer Asked for a Bribe
	
```{r}
 
exc2 <- ddply(lapop.trends,~year,summarize,x=sum(!is.na(exc2) & exc2==1),
            n=sum(!is.na(exc2) & (exc2==0 | exc2==1)))

AmericasBarometer::trend_plot(exc2,
                              'Asked for a bribe by a police officer?',
                              '   ')
``` 

###	exc20	Soldier Requested a Bribe
	
```{r}
#table(lapop.trends$exc20, useNA= "ifany") 
exc20  <- ddply(lapop.trends,~year,summarize,x=sum(!is.na(exc20 ) & exc20 ==1),
            n=sum(!is.na(exc20) & (exc20==0 | exc20==1)))

AmericasBarometer::trend_plot(exc20,
                              'Soldier Requested a Bribe',
                              '  ')
```	

### exc6	Government Employee Asked for a Bribe
	
```{r}
#table(lapop.trends$exc6, useNA= "ifany") 
exc6  <- ddply(lapop.trends,~year,summarize,x=sum(!is.na(exc6 ) & exc6 ==1),
            n=sum(!is.na(exc6) & (exc6==0 | exc6==1)))

AmericasBarometer::trend_plot(exc6,
                              'Government Employee Asked for a Bribe',
                              '  ')
```	

### exc7	Perception of Public Corruption among Public Officials
	
```{r}
# table(lapop.trends$exc7, lapop.trends$wave, useNA= "ifany")  
exc7  <- ddply(lapop.trends,~year,summarize,x=sum(!is.na(exc7 ) & exc7 ==1),
            n=sum(!is.na(exc7) & (exc7==2 | exc7==1)))

AmericasBarometer::trend_plot(exc7,
                              'Perception of Public Corruption among Public Officials',
                              '  ')
```	

### exc7new	Amount of Corruption among Politicians
	
```{r}
#table(lapop.trends$exc7new	,lapop.trends$wave	, useNA= "ifany") 
exc7new <- ddply(lapop.trends,~year,summarize,
             Nonne=sum(exc7new==1,na.rm=TRUE) / sum(exc7new < 6,na.rm=TRUE),
             Less_than_Half=sum(exc7new==2,na.rm=TRUE) / sum(exc7new < 6,na.rm=TRUE),
             Half=sum(exc7new==3,na.rm=TRUE) / sum(exc7new < 6,na.rm=TRUE),
             More_then_half=sum(exc7new==4,na.rm=TRUE) / sum(exc7new < 6,na.rm=TRUE),
             All=sum(exc7new==5,na.rm=TRUE) / sum(exc7new < 6,na.rm=TRUE))

AmericasBarometer::multi_lines(exc7new,
                              'Amount of Corruption among Politicians',
                              '  ')
```	

 
	
	
## Crime, Insecurity, and Policy Preferences	

### aoj11	Perception of Neighborhood Insecurity
	
```{r}
#table(lapop.trends$aoj11, useNA= "ifany") 
aoj11 <- ddply(lapop.trends,~year,summarize,
             Very_secured =sum(aoj11==1,na.rm=TRUE) / sum(aoj11 < 5,na.rm=TRUE),
             Secured =sum(aoj11==2,na.rm=TRUE) / sum(aoj11 < 5,na.rm=TRUE),
             Insecured =sum(aoj11==3,na.rm=TRUE) / sum(aoj11 < 5,na.rm=TRUE),
             Very_Insecured =sum(aoj11==4,na.rm=TRUE) / sum(aoj11 < 5,na.rm=TRUE))

AmericasBarometer::multi_lines(aoj11,
                              'Perception of Neighborhood Insecurity',
                              '  ')
```	

###	aoj12	Confidence that Judiciary will Punish the Guilty
	
```{r}
#table(lapop.trends$aoj12, useNA= "ifany") 
aoj12 <- ddply(lapop.trends,~year,summarize,
             Very =sum(aoj12==1,na.rm=TRUE) / sum(aoj12 < 5,na.rm=TRUE),
             Somewhat  =sum(aoj12==2,na.rm=TRUE) / sum(aoj12 < 5,na.rm=TRUE),
             Little =sum(aoj12==3,na.rm=TRUE) / sum(aoj12 < 5,na.rm=TRUE),
             Not =sum(aoj12==4,na.rm=TRUE) / sum(aoj12 < 5,na.rm=TRUE))

AmericasBarometer::multi_lines(aoj12,
                              'Confidence that Judiciary will Punish the Guilty',
                              '  ')
```	

###	vic1ext	Victim of Crime
	
```{r}
# table(lapop.trends$vic1ext, lapop.trends$wave, useNA= "ifany") 
vic1ext  <- ddply(lapop.trends,~year,summarize,x=sum(!is.na(vic1ext ) & vic1ext ==1),
            n=sum(!is.na(vic1ext) & (vic1ext==2 | vic1ext==1)))

AmericasBarometer::trend_plot(vic1ext,
                              'Victim of Crime',
                              '  ')
```	

###	vic1exta	Victim of Crime in the Last 12 Months (Frequency)
	
```{r}
# table(lapop.trends$vic1exta, lapop.trends$wave, useNA= "ifany")
exc7  <- ddply(lapop.trends,~year,summarize,x=sum(!is.na(exc7 ) & exc7 ==1),
            n=sum(!is.na(exc7) & (exc7==2 | exc7==1)))

AmericasBarometer::trend_plot(exc7,
                              'Victim of Crime in the Last 12 Months (Frequency)',
                              '  ')
```	


	
## Evaluation of Institutions, Political System and Other Persons	


### a4	Most Important Problem
	
```{r}
# table(lapop.trends$a4, lapop.trends$wave, useNA= "ifany")  
exc7  <- ddply(lapop.trends,~year,summarize,x=sum(!is.na(exc7 ) & exc7 ==1),
            n=sum(!is.na(exc7) & (exc7==2 | exc7==1)))

AmericasBarometer::trend_plot(exc7,
                              'Perception of Public Corruption among Public Officials',
                              '  ')
```	

###	b1	Courts Guarantee Fair Trial
	
```{r}
# table(lapop.trends$b1, lapop.trends$wave, useNA= "ifany")  
exc7  <- ddply(lapop.trends,~year,summarize,x=sum(!is.na(exc7 ) & exc7 ==1),
            n=sum(!is.na(exc7) & (exc7==2 | exc7==1)))

AmericasBarometer::trend_plot(exc7,
                              'Perception of Public Corruption among Public Officials',
                              '  ')
```	

###	b12	Trust in Armed Forces
	
```{r}
# table(lapop.trends$b12, lapop.trends$wave, useNA= "ifany")  
exc7  <- ddply(lapop.trends,~year,summarize,x=sum(!is.na(exc7 ) & exc7 ==1),
            n=sum(!is.na(exc7) & (exc7==2 | exc7==1)))

AmericasBarometer::trend_plot(exc7,
                              'Perception of Public Corruption among Public Officials',
                              '  ')
```	

###	b13	Trust in the National Legislature
	
```{r}
# table(lapop.trends$exc7, lapop.trends$wave, useNA= "ifany")  
exc7  <- ddply(lapop.trends,~year,summarize,x=sum(!is.na(exc7 ) & exc7 ==1),
            n=sum(!is.na(exc7) & (exc7==2 | exc7==1)))

AmericasBarometer::trend_plot(exc7,
                              'Perception of Public Corruption among Public Officials',
                              '  ')
```	

###	b18	Trust in National Police
	
```{r}
# table(lapop.trends$exc7, lapop.trends$wave, useNA= "ifany")  
exc7  <- ddply(lapop.trends,~year,summarize,x=sum(!is.na(exc7 ) & exc7 ==1),
            n=sum(!is.na(exc7) & (exc7==2 | exc7==1)))

AmericasBarometer::trend_plot(exc7,
                              'Perception of Public Corruption among Public Officials',
                              '  ')
```	

###	b2	Respect for Political Institutions
	
```{r}
# table(lapop.trends$exc7, lapop.trends$wave, useNA= "ifany")  
exc7  <- ddply(lapop.trends,~year,summarize,x=sum(!is.na(exc7 ) & exc7 ==1),
            n=sum(!is.na(exc7) & (exc7==2 | exc7==1)))

AmericasBarometer::trend_plot(exc7,
                              'Perception of Public Corruption among Public Officials',
                              '  ')
```	

###	b20	Trust in Catholic Church
	
```{r}
# table(lapop.trends$exc7, lapop.trends$wave, useNA= "ifany")  
exc7  <- ddply(lapop.trends,~year,summarize,x=sum(!is.na(exc7 ) & exc7 ==1),
            n=sum(!is.na(exc7) & (exc7==2 | exc7==1)))

AmericasBarometer::trend_plot(exc7,
                              'Perception of Public Corruption among Public Officials',
                              '  ')
```	

###	b20a	Trust in Evangelical/Protestant Church
	
```{r}
# table(lapop.trends$exc7, lapop.trends$wave, useNA= "ifany")  
exc7  <- ddply(lapop.trends,~year,summarize,x=sum(!is.na(exc7 ) & exc7 ==1),
            n=sum(!is.na(exc7) & (exc7==2 | exc7==1)))

AmericasBarometer::trend_plot(exc7,
                              'Perception of Public Corruption among Public Officials',
                              '  ')
```	

###	b21	Trust in Political Parties
	
```{r}
# table(lapop.trends$exc7, lapop.trends$wave, useNA= "ifany")  
exc7  <- ddply(lapop.trends,~year,summarize,x=sum(!is.na(exc7 ) & exc7 ==1),
            n=sum(!is.na(exc7) & (exc7==2 | exc7==1)))

AmericasBarometer::trend_plot(exc7,
                              'Perception of Public Corruption among Public Officials',
                              '  ')
```	

###	b21a	Trust in Executive
	
```{r}
# table(lapop.trends$exc7, lapop.trends$wave, useNA= "ifany")  
exc7  <- ddply(lapop.trends,~year,summarize,x=sum(!is.na(exc7 ) & exc7 ==1),
            n=sum(!is.na(exc7) & (exc7==2 | exc7==1)))

AmericasBarometer::trend_plot(exc7,
                              'Perception of Public Corruption among Public Officials',
                              '  ')
```	

###	b3	Basic Rights are Protected
	
```{r}
# table(lapop.trends$exc7, lapop.trends$wave, useNA= "ifany")  
exc7  <- ddply(lapop.trends,~year,summarize,x=sum(!is.na(exc7 ) & exc7 ==1),
            n=sum(!is.na(exc7) & (exc7==2 | exc7==1)))

AmericasBarometer::trend_plot(exc7,
                              'Perception of Public Corruption among Public Officials',
                              '  ')
```	

###	b31	Trust in the Supreme Court of Justice
	
```{r}
# table(lapop.trends$exc7, lapop.trends$wave, useNA= "ifany")  
exc7  <- ddply(lapop.trends,~year,summarize,x=sum(!is.na(exc7 ) & exc7 ==1),
            n=sum(!is.na(exc7) & (exc7==2 | exc7==1)))

AmericasBarometer::trend_plot(exc7,
                              'Perception of Public Corruption among Public Officials',
                              '  ')
```	

###	b32	Trust in Local Government
	
```{r}
# table(lapop.trends$exc7, lapop.trends$wave, useNA= "ifany")  
exc7  <- ddply(lapop.trends,~year,summarize,x=sum(!is.na(exc7 ) & exc7 ==1),
            n=sum(!is.na(exc7) & (exc7==2 | exc7==1)))

AmericasBarometer::trend_plot(exc7,
                              'Perception of Public Corruption among Public Officials',
                              '  ')
```	

###	b37	Trust in the Media
	
```{r}
# table(lapop.trends$exc7, lapop.trends$wave, useNA= "ifany")  
exc7  <- ddply(lapop.trends,~year,summarize,x=sum(!is.na(exc7 ) & exc7 ==1),
            n=sum(!is.na(exc7) & (exc7==2 | exc7==1)))

AmericasBarometer::trend_plot(exc7,
                              'Perception of Public Corruption among Public Officials',
                              '  ')
```	

###	b4	Pride in Political System
	
```{r}
# table(lapop.trends$exc7, lapop.trends$wave, useNA= "ifany")  
exc7  <- ddply(lapop.trends,~year,summarize,x=sum(!is.na(exc7 ) & exc7 ==1),
            n=sum(!is.na(exc7) & (exc7==2 | exc7==1)))

AmericasBarometer::trend_plot(exc7,
                              'Perception of Public Corruption among Public Officials',
                              '  ')
```	

###	b47a	Trust in Elections
	
```{r}
# table(lapop.trends$exc7, lapop.trends$wave, useNA= "ifany")  
exc7  <- ddply(lapop.trends,~year,summarize,x=sum(!is.na(exc7 ) & exc7 ==1),
            n=sum(!is.na(exc7) & (exc7==2 | exc7==1)))

AmericasBarometer::trend_plot(exc7,
                              'Perception of Public Corruption among Public Officials',
                              '  ')
```	

###	b6	People Should Support the Political System
	
```{r}
# table(lapop.trends$exc7, lapop.trends$wave, useNA= "ifany")  
exc7  <- ddply(lapop.trends,~year,summarize,x=sum(!is.na(exc7 ) & exc7 ==1),
            n=sum(!is.na(exc7) & (exc7==2 | exc7==1)))

AmericasBarometer::trend_plot(exc7,
                              'Perception of Public Corruption among Public Officials',
                              '  ')
```	

###	eff1	Leaders Are Interested in What People Think
	
```{r}
# table(lapop.trends$exc7, lapop.trends$wave, useNA= "ifany")  
exc7  <- ddply(lapop.trends,~year,summarize,x=sum(!is.na(exc7 ) & exc7 ==1),
            n=sum(!is.na(exc7) & (exc7==2 | exc7==1)))

AmericasBarometer::trend_plot(exc7,
                              'Perception of Public Corruption among Public Officials',
                              '  ')
```	

###	eff2	Understands Important Political Issues
	
```{r}
# table(lapop.trends$eff2, lapop.trends$wave, useNA= "ifany")  
exc7  <- ddply(lapop.trends,~year,summarize,x=sum(!is.na(exc7 ) & exc7 ==1),
            n=sum(!is.na(exc7) & (exc7==2 | exc7==1)))

AmericasBarometer::trend_plot(exc7,
                              'Perception of Public Corruption among Public Officials',
                              '  ')
```	

###	it1	Interpersonal Trust
	
```{r}
# table(lapop.trends$exc7, lapop.trends$wave, useNA= "ifany")  
exc7  <- ddply(lapop.trends,~year,summarize,x=sum(!is.na(exc7 ) & exc7 ==1),
            n=sum(!is.na(exc7) & (exc7==2 | exc7==1)))

AmericasBarometer::trend_plot(exc7,
                              'Perception of Public Corruption among Public Officials',
                              '  ')
```	

###	lib1	Level of Freedom of the Press Today
	
```{r}
# table(lapop.trends$exc7, lapop.trends$wave, useNA= "ifany")  
exc7  <- ddply(lapop.trends,~year,summarize,x=sum(!is.na(exc7 ) & exc7 ==1),
            n=sum(!is.na(exc7) & (exc7==2 | exc7==1)))

AmericasBarometer::trend_plot(exc7,
                              'Perception of Public Corruption among Public Officials',
                              '  ')
```	

###	lib2c	Level of Freedom to Express Political Opinions without Fear
	
```{r}
# table(lapop.trends$exc7, lapop.trends$wave, useNA= "ifany")  
exc7  <- ddply(lapop.trends,~year,summarize,x=sum(!is.na(exc7 ) & exc7 ==1),
            n=sum(!is.na(exc7) & (exc7==2 | exc7==1)))

AmericasBarometer::trend_plot(exc7,
                              'Perception of Public Corruption among Public Officials',
                              '  ')
```	

###	m1	Executive Job Approval
	
```{r}
# table(lapop.trends$exc7, lapop.trends$wave, useNA= "ifany")  
exc7  <- ddply(lapop.trends,~year,summarize,x=sum(!is.na(exc7 ) & exc7 ==1),
            n=sum(!is.na(exc7) & (exc7==2 | exc7==1)))

AmericasBarometer::trend_plot(exc7,
                              'Perception of Public Corruption among Public Officials',
                              '  ')
```	

###	Military coup is justified if crime is high

Whatever happened in 2008, it also coincided with a sudden rise in the number of people who are open to the idea of a military coup.
```{r}
jc10 <- ddply(lapop.trends,~year,summarize,x=sum(!is.na(jc10) & jc10==2),
            n=sum(!is.na(jc10) & (jc10==1 | jc10==2)))
AmericasBarometer::trend_plot(jc10,
                              'Military coup is justified if crime is high',
                              'jc10')
```
	
## Foreign Relations	

###	mil10a	Trustworthiness of Chinese Government
	
```{r}
# table(lapop.trends$exc7, lapop.trends$wave, useNA= "ifany")  
exc7  <- ddply(lapop.trends,~year,summarize,x=sum(!is.na(exc7 ) & exc7 ==1),
            n=sum(!is.na(exc7) & (exc7==2 | exc7==1)))

AmericasBarometer::trend_plot(exc7,
                              'Perception of Public Corruption among Public Officials',
                              '  ')
```	

###	mil10e	Trustworthiness of US Government
	
```{r}
# table(lapop.trends$exc7, lapop.trends$wave, useNA= "ifany")  
exc7  <- ddply(lapop.trends,~year,summarize,x=sum(!is.na(exc7 ) & exc7 ==1),
            n=sum(!is.na(exc7) & (exc7==2 | exc7==1)))

AmericasBarometer::trend_plot(exc7,
                              'Perception of Public Corruption among Public Officials',
                              '  ')
```	



## Gender	


	
```{r}
#v58	Children Suffer with Working Mothers
# table(lapop.trends$v58, lapop.trends$wave, useNA= "ifany")  
```	

###	vb50	Men Are Better Political Leaders than Women
	
```{r}
# table(lapop.trends$vb50, lapop.trends$wave, useNA= "ifany")  
exc7  <- ddply(lapop.trends,~year,summarize,x=sum(!is.na(exc7 ) & exc7 ==1),
            n=sum(!is.na(exc7) & (exc7==2 | exc7==1)))

AmericasBarometer::trend_plot(exc7,
                              'Perception of Public Corruption among Public Officials',
                              '  ')
```	

###	vb51	Men or Women: More Corrupt as Politicians
	
```{r}
# table(lapop.trends$vb51, lapop.trends$wave, useNA= "ifany")  
exc7  <- ddply(lapop.trends,~year,summarize,x=sum(!is.na(exc7 ) & exc7 ==1),
            n=sum(!is.na(exc7) & (exc7==2 | exc7==1)))

AmericasBarometer::trend_plot(exc7,
                              'Perception of Public Corruption among Public Officials',
                              '  ')
```	

###	vb52	Men or Women: Better Manager of Economy
	
```{r}
# table(lapop.trends$exc7, lapop.trends$wave, useNA= "ifany")  
exc7  <- ddply(lapop.trends,~year,summarize,x=sum(!is.na(exc7 ) & exc7 ==1),
            n=sum(!is.na(exc7) & (exc7==2 | exc7==1)))

AmericasBarometer::trend_plot(exc7,
                              'Perception of Public Corruption among Public Officials',
                              '  ')
```	

###	w14a	Abortion Justified
	
```{r}
# table(lapop.trends$exc7, lapop.trends$wave, useNA= "ifany")  
exc7  <- ddply(lapop.trends,~year,summarize,x=sum(!is.na(exc7 ) & exc7 ==1),
            n=sum(!is.na(exc7) & (exc7==2 | exc7==1)))

AmericasBarometer::trend_plot(exc7,
                              'Perception of Public Corruption among Public Officials',
                              '  ')
```	


	
## Local Government	
###	  infrax	Police Response Time
	
```{r}
# table(lapop.trends$exc7, lapop.trends$wave, useNA= "ifany")  
exc7  <- ddply(lapop.trends,~year,summarize,x=sum(!is.na(exc7 ) & exc7 ==1),
            n=sum(!is.na(exc7) & (exc7==2 | exc7==1)))

AmericasBarometer::trend_plot(exc7,
                              'Perception of Public Corruption among Public Officials',
                              '  ')
```	

###	np1	Attendance at Municipal Meeting
	
```{r}
# table(lapop.trends$exc7, lapop.trends$wave, useNA= "ifany")  
exc7  <- ddply(lapop.trends,~year,summarize,x=sum(!is.na(exc7 ) & exc7 ==1),
            n=sum(!is.na(exc7) & (exc7==2 | exc7==1)))

AmericasBarometer::trend_plot(exc7,
                              'Perception of Public Corruption among Public Officials',
                              '  ')
```	

###	sd2new2	Satisfaction with Roads
	
```{r}
# table(lapop.trends$exc7, lapop.trends$wave, useNA= "ifany")  
exc7  <- ddply(lapop.trends,~year,summarize,x=sum(!is.na(exc7 ) & exc7 ==1),
            n=sum(!is.na(exc7) & (exc7==2 | exc7==1)))

AmericasBarometer::trend_plot(exc7,
                              'Perception of Public Corruption among Public Officials',
                              '  ')
```	

###	sd3new2	Satisfaction with Public Schools
	
```{r}
# table(lapop.trends$exc7, lapop.trends$wave, useNA= "ifany")  
exc7  <- ddply(lapop.trends,~year,summarize,x=sum(!is.na(exc7 ) & exc7 ==1),
            n=sum(!is.na(exc7) & (exc7==2 | exc7==1)))

AmericasBarometer::trend_plot(exc7,
                              'Perception of Public Corruption among Public Officials',
                              '  ')
```	

###	sd6new2	Satisfaction with Public Medical and Health Services
	
```{r}
#table(lapop.trends$sd6new2, useNA= "ifany") 
exc7  <- ddply(lapop.trends,~year,summarize,x=sum(!is.na(exc7 ) & exc7 ==1),
            n=sum(!is.na(exc7) & (exc7==2 | exc7==1)))

AmericasBarometer::trend_plot(exc7,
                              'Perception of Public Corruption among Public Officials',
                              '  ')
```	


## Possession

### Refrigerator in home

Refrigerators are on the rise.
```{r}
r3 <- ddply(lapop.trends,~year,summarize,x=sum(!is.na(r3) & r3==1),
            n=sum(!is.na(r3) & (r3==0 | r3==1)))
AmericasBarometer::trend_plot(r3,
                              'Refrigerator in home',
                               '')
```

### Landline telephone in home
Landlines are on their way out.
```{r}
r4 <- ddply(lapop.trends,~year,summarize,x=sum(!is.na(r4) & r4==1),
             n=sum(!is.na(r4) & (r4==0 | r4==1)))
AmericasBarometer::trend_plot(r4,
                              'Landline telephone in home',
                              '')
```

### Cellular telephone in home



The cell phone market looks like it saturated in about 2010.

```{r}
r4a <- ddply(lapop.trends,~year,summarize,x=sum(!is.na(r4a) & r4a==1),
             n=sum(!is.na(r4a) & (r4a==0 | r4a==1)))
AmericasBarometer::trend_plot(r4a,
                              'Cellular telephone in home',
                              '')
```

 

## Education 
Education seems to be improving -- there has been a decrease in people with less than five years of education, while the fraction of elementary-school graduates is holding fairly steady and the fraction of high-school graduates is increasing. It would be easy enough to subset this analysis to individual countries or regions.
```{r}
ed <- lapop.trends[lapop.trends$year %in% c(2004,2008,2014,2016,2018),c('ed','year')]
ed$year <- as.character(ed$year)
col3 <- brewer.pal(3,"Dark2")
ed <- na.omit(ed)


ggplot(ed,aes(x=ed,group=year,color=year)) +
  geom_density(size=2) +
  scale_color_brewer(type="qual") +
  scale_x_continuous(limits=c(0,max(lapop.trends$ed,na.rm=TRUE))) +
  theme_classic() +
  xlab("Years of education")+
  theme(text=element_text(size=20),
        axis.title.y=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank(),
        axis.line.y=element_blank()) 
```













